Estimation below the detection limit

### 2018-04-12 TEX/draft2.tex
Pushing more updates to the paper about 1D XP

For the purpose of this paper, we need the first three (aperiodic) autocorrelation functions. The first-order autocorelation is the mean of the signals. For  
$z\in\RL$ and $k\geq 2$, the higher order autocorrelations are defined by
\begin{equation} \label{eq:ac_general}
a_z^k[\ell_1,\ldots,\ell_{k-1}] =  \sum_{i=0}^{L-1-\max\{\ell_1,\ldots,\ell_{k-1}\}} z[i]z[i+\ell_1]\ldots z[i+\ell_{k-1}].
\end{equation}
Explicitly, the first three autocorrelations are
\begin{align} \label{eq:autocorrelations}
a_z^1 &= \sum_{i=0}^{L-1} z[i], \nonumber\\
a_z^2[\ell] &= \sum_{i=0}^{L-1-\ell} z[i]z[i+\ell], \nonumber\\
a_z^3[\ell_1,\ell_2] &= \sum_{i=0}^{L-1-\max\{\ell_1,\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. 
\end{align}
Note that the autocorrelation functions are symmetric so that $a_z^2[\ell] = a_z^2[-\ell]$ and $a_z^3[\ell_1,\ell_2] = a_z^3[-\ell_1,-\ell_2]$. \TODO{We also have $a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]$ -- is that intended? Do we really have symmetry with the negative signs? I think we have $(\ell_1, \ell_2) \sim (-\ell_1, \ell_2-\ell_1)$; to be checked, but I don't think this is correct;}
Additionally, if the moments of the signal depend only on the difference between the indices (Toeplitz structure), then they are equivalent to the autocorrelation functions.

A one-dimensional signal is determined uniquely and stably by its third-order auto-correlation as proven in the following simple proposition.
\begin{proposition} \label{prop:uniqueness}
	Let $z\in\RL$ and suppose that $z[0]$ and $z[L-1]$ are nonzeros. Then:
	\begin{itemize}
		\item \textbf{Uniqueness:} 	 $z$  is determined uniquely from  $a_z^2$ and $a_z^3$.
        \item \textbf{Finite sensitivity:} 	Suppose that  $a_z^3[k,L-1]$ can be estimated up to  perturbation $\upsilon$ and let $\hat{z}[k]$ be an estimator of $z[k]$. If $\vert z[0]z[L-1]\vert \geq \delta>0$, then $\vert \hat{z}[k] - z[k]\vert\leq \frac{\vert \upsilon\vert }{\delta}$. 
	\end{itemize}
	\begin{proof}
		By assumption $a_z^2[L-1] = z[0]z[L-1]\neq 0$.
		Then, the uniqueness results, for all $k=0,\ldots L-1$,  follows from:
		\begin{equation*}
		a_z^3[k,L-1] = z[0]z[k]z[L-1].
		\end{equation*}
		If we measure $\tilde{a_z^3}[k,L-1] = a_z^3[k,L-1]+\upsilon$, then 
		\begin{equation*}
		\hat{z}[k] = \frac{\tilde{a_z^3}[k,L-1]}{a_z^2[L-1]} = z[k]+\frac{\upsilon}{a_z^2[L-1]} \quad \Rightarrow \quad \vert \hat{z}[k] - {z}[k]\vert \leq \frac{\vert\upsilon\vert}{\delta}.
		\end{equation*} 
	\end{proof}
\end{proposition}

@@

The invertibility of the autocorrelations for $K>1$ was explored  for the related case of  periodic autocorrelation functions. Recall that the  periodic autocorrelations are defined similarly to~\eqref{eq:ac_general}, with two differences: the sum goes to $L-1$ and all indices should be taken modulo $L$.
If $z[n]=0$ for $n=L/2,\ldots L-1$ (``zero-padded''), then the periodic and aperiodic autocorrelation coincide.
In~\cite{bandeira2017estimation}, it was shown that a mix of $K$ third-order periodic autocorrelations determine  a finite list of $K$ generic signals if $L/6\gtrsim K$. Empirical evidences hints that this finite list includes only group symmetries, at least as long as $K\leq\sqrt{L}$~\cite{boumal2017heterogeneous}.

@@

When $K>1$, it is impossible to estimate the autocorrelation of each signal directly from the data. However similarly to~\eqref{eq:data_ac_k1}, one can estimate the mixture of the autocorrelations. 
As before, we consider  the asymptomatic regime where $M_1,\ldots,M_K,N\to\infty$, while preserving fixed ratios
\begin{equation}
\gamma_i = \frac{M_i L}{N}, \quad \gamma = \sum_{i=1}^K\gamma_i.
\end{equation}
If the support $s$ satisfies the spacing constraint~\eqref{eq:spacing}, then:
\begin{align}
\lim_{N\to\infty} a_y^1 &= \sum_{i=1}^K\gamma_i a_{x_i}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] &= \sum_{i=1}^K\gamma_i a_{x_i}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] &= \sum_{i=1}^K\gamma_i a_{x_i}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{i=1}^K\gamma_i a_{x_i}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
This relation is proven in  Appendix~\ref{sec:autocorrelation_computation}.

@@

Given the observation $y$, we proceed to compute the moments. The first-order moment is straightforward. For second-order moments, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to estimate the moments. Likewise, for third-order moments, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant moments for our purpose \TODO{Do we want to include a figure to explain why that is?}, and we exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid biased elements---there are $\frac{(L-1)(L-2)}{2}$ such moments. As a result, it is unnecessary to estimate $\sigma$. There are
\begin{align*}
	1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
moments in total. In practice, these are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data.

>>>>

For the purpose of this paper, we need the first three (aperiodic) autocorrelation functions. The first-order autocorelation is the mean of the signals. For  
$z\in\RL$ and $k\geq 2$, the autocorrelation of order $k$ is defined for any integer shifts $\ell_1, \ldots, \ell_{k-1}$ by
\begin{align}
	a_z^k[\ell_1,\ldots,\ell_{k-1}]  & = \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\ldots z[i+\ell_{k-1}],
	\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the bounds $0, \ldots, L-1$ is zero-padded, as usual.
Explicitly, the first three autocorrelations are
\begin{align} 
	a_z^1 & = \sum_{i=0}^{L-1} z[i], \nonumber\\
	a_z^2[\ell] & = \sum_{i = \max\{0, -\ell\}}^{L-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
	a_z^3[\ell_1,\ell_2] & = \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{L-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
Note that the autocorrelation functions are symmetric so that $a_z^2[\ell] = a_z^2[-\ell]$ and $a_z^3[\ell_1,\ell_2] = a_z^3[-\ell_1,-\ell_2]$. \TODO{We also have $a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]$ -- is that intended? Do we really have symmetry with the negative signs? I think we have $(\ell_1, \ell_2) \sim (-\ell_1, \ell_2-\ell_1)$; to be checked, but I don't think this is correct;}
Additionally, if the moments of the signal depend only on the difference between the indices (Toeplitz structure), then they are equivalent to the autocorrelation functions.

A one-dimensional signal is determined uniquely and stably by its third-order auto-correlation as proven in the following simple proposition.
\begin{proposition} \label{prop:uniqueness}
	Let $z\in\RL$ and suppose that $z[0]$ and $z[L-1]$ are nonzero. Then:
	\begin{itemize}
		\item \textbf{Uniqueness:} 	 $z$  is determined uniquely from  $a_z^2$ and $a_z^3$.
        \item \textbf{Finite sensitivity:} 	Suppose that  $a_z^3[k,L-1]$ can be estimated up to  perturbation $\upsilon$ and let $\hat{z}[k]$ be an estimator of $z[k]$. If $\vert z[0]z[L-1]\vert \geq \delta>0$, then $\vert \hat{z}[k] - z[k]\vert\leq \frac{\vert \upsilon\vert }{\delta}$. \TODO{That statement is incomplete: you mean there exists an estimator that guarantees this. (Since the estimator is very simple, we may want to show it explicitly in the proposition statement, and comment afterwards that of course this is still fairly sensitive, and we can do better.)}
	\end{itemize}
	\begin{proof}
		By assumption $a_z^2[L-1] = z[0]z[L-1]\neq 0$.
		Then, the uniqueness results, for all $k=0,\ldots L-1$,  follows from:
		\begin{equation*}
		a_z^3[k,L-1] = z[0]z[k]z[L-1].
		\end{equation*}
		If we measure $\tilde{a_z^3}[k,L-1] = a_z^3[k,L-1]+\upsilon$, then 
		\begin{equation*}
		\hat{z}[k] = \frac{\tilde{a_z^3}[k,L-1]}{a_z^2[L-1]} = z[k]+\frac{\upsilon}{a_z^2[L-1]} \quad \Rightarrow \quad \vert \hat{z}[k] - {z}[k]\vert \leq \frac{\vert\upsilon\vert}{\delta}.
		\end{equation*} 
	\end{proof}
\end{proposition}

@@

\TODO{I would remove most if this paragraph; we can have a similar discussion but for our case (after having discussing which moments we keep); make the statement, and finish with a one-sentence reference to the MRA paper.}
The invertibility of the autocorrelations for $K>1$ was explored  for the related case of  periodic autocorrelation functions. Recall that the  periodic autocorrelations are defined similarly to~\eqref{eq:ac_general}, with two differences: the sum goes to $L-1$ and all indices should be taken modulo $L$.
If $z[n]=0$ for $n=L/2,\ldots L-1$ (``zero-padded''), then the periodic and aperiodic autocorrelation coincide.
In~\cite{bandeira2017estimation}, it was shown that a mix of $K$ third-order periodic autocorrelations determine  a finite list of $K$ generic signals if $L/6\gtrsim K$. Empirical evidences hints that this finite list includes only group symmetries, at least as long as $K\leq\sqrt{L}$~\cite{boumal2017heterogeneous}.

@@

When $K>1$, it is impossible \TODO{That's not true: we can do our procedure, which gives estimate of the signals, and moments of that are estimates of the moments of the individual signals. Whether one calls this 'direct' or not is subjective. I'd remove that sentence.)} to estimate the autocorrelation of each signal directly from the data. However similarly to~\eqref{eq:data_ac_k1}, one can estimate the mixture of the autocorrelations. 
As before, we consider  the asymptomatic regime where $M_1,\ldots,M_K,N\to\infty$, while preserving fixed ratios
\begin{align}
	\gamma_k = \frac{M_k L}{N}, \quad \gamma = \sum_{k=1}^K\gamma_k.
\end{align}
If the support $s$ satisfies the spacing constraint~\eqref{eq:spacing}, then:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
This relation is proven in  Appendix~\ref{sec:autocorrelation_computation}.

@@

Given the observation $y$, we proceed to compute the moments. The first-order moment is straightforward. For second-order moments, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to estimate the moments. Likewise, for third-order moments, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant moments for our purpose \TODO{Do we want to include a figure to explain why that is?}, and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid biased elements---there are $\frac{(L-1)(L-2)}{2}$ remaining moments. As a result, it is unnecessary to estimate $\sigma$. We have
\begin{align*}
	1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
moments in total. In practice, these are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
\begin{multline}
	\min_{\substack{\hat x_1, \ldots, \hat x_K \in \reals^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2.
	\label{eq:optim1D}
\end{multline}
where $W \geq L$ is the length of the sought signals and \TODO{explain $w_i$'s}. Setting $W = L$ (as is a priori desired) is problematic as the above optimization problems appears to have numerous poor local optimizers. Since we can only initialize randomly at first, this approach would often fail in practice. Alternatively, we first run the optimization with $W = 2L-1$. This problem appears to have fewer poor local optima, perhaps as a result of the additional degrees of freedom which allow for more escape directions. As we would hope to find that these estimated signals are supported on $L$ contiguous entries, we extract from each one a subsignal of length $L$ (with cyclic indexing) that has largest $\ell_2$-norm. This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this is reliable for a wide range of experimental parameters.
